Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data

### 2025-01-12 paper/paper.tex
Expanded intro

\begin{abstract}
Auditing AI systems often requires sensitive data, raising security and privacy concerns. To address this, we propose a framework that leverages differentially private synthetic data for fairness evaluations. This approach enables auditors to assess AI systems without exposing sensitive information. Through experiments on real datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. Our results show that synthetic data accurately reflect the fairness properties of real data to a reasonable extent.
\end{abstract}

@@

% TODO: add to intro to fill 14 pgs

@@

Fairness in machine learning has become an important topic as AI systems become widely used. Biased AI systems could result in amplified unfairness. Ensuring fairness in AI systems is crucial to prevent these biases from being exacerbated by AI systems. One approach is to audit the fairness of existing AI systems. Quantitative fairness auditing is a process of evaluating the fairness of AI systems with a range of fairness metrics. Many metrics have been proposed to help evaluate the fairness of AI systems.

Auditing can evaluate the fairness of AI systems, sometimes revealing cases of unfairness in them. For example, investigative journalists at ProPublica found that the COMPAS system, which is used to predict recidivism in the criminal justice system in the United States, is biased against certain demographic groups. This highlights the need for fairness auditing conducted by third parties.

@@

To conduct audits, auditors may rely on access to real datasets containing sensitive information, raising significant security and privacy concerns. This reliance invites data security attacks on auditors, making them targets of unauthorized access. Analysis on real data runs the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Various technologies have been developed to mitigate these risks, such as differential privacy, where the idea is that whether or not any individual is in a dataset has a limited impact.

To mitigate these risks, the use of synthetic data has been proposed. Synthetic data are fake data that is artificially generated from real data. The principle of this technology is to preserve some statistical properties of the real data while protecting the privacy of the individuals in the real data. Among the generation techniques, differentially private synthetic data offers a way to investigate the data properties under differential privacy guarantees.

@@

We propose a novel auditing framework that leverages differentially private synthetic data to evaluate the fairness of AI systems. Our framework lets the auditors assess the fairness of AI systems without exposing sensitive information. Differentially private synthetic data is a technique that generates synthetic data from real data that preserves some statistical properties of the real data while ensuring differential privacy. This technology enables auditing without exposing sensitive information. However, synthetic data also introduces new challenges. Whether or not it accurately preserves the fairness properties of the real data is a critical question.

@@

@@

>>>>

\begin{abstract}
Fairness auditing of AI systems can identify and quantify biases. However, traditional auditing using real-world data raises security and privacy concerns. It exposes auditors to security risks as they become custodians of sensitive information and targets for cyberattacks. Privacy risks arise even without direct breaches, as data analyses can inadvertently expose confidential information. To address these, we propose a framework that leverages differentially private synthetic data to audit the fairness of AI systems. By applying privacy-preserving mechanisms, it generates synthetic data that mirrors the statistical properties of the original dataset while ensuring individual privacy. This method balances the goal of rigorous fairness auditing and the need for strong privacy protections. Through experiments on real datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. By analyzing the alignment and discrepancies between these metrics, we assess the capacity of synthetic data to preserve the fairness properties of real data. Our results demonstrate the framework's ability to enable meaningful fairness evaluations while safeguarding sensitive information, proving its applicability across critical and sensitive domains.
% Auditing AI systems often requires sensitive data, raising security and privacy concerns. To address this, we propose a framework that leverages differentially private synthetic data for fairness evaluations. This approach enables auditors to assess AI systems without exposing sensitive information. Through experiments on real datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. Our results show that synthetic data accurately reflect the fairness properties of real data to a reasonable extent.
\end{abstract}

@@

@@

\subsection{Fairness in AI}

Fairness in machine learning has become an increasingly important topic as AI systems are widely used across various sectors, including healthcare and criminal justice. These systems can significantly influence outcomes that impact individuals and communities.

However, AI systems can exhibit bias, which may lead to increased unfairness. Biased AI systems have the potential to reinforce historical injustices, erode trust in technology, and exacerbate inequality. Therefore, ensuring fairness in AI systems is crucial to prevent the perpetuation of these biases.

For instance, investigative journalists at ProPublica uncovered bias in the COMPAS system, which is used in the United States criminal justice system to predict recidivism. Their analysis revealed that the COMPAS system disproportionately affects certain demographic groups.

These findings have raised significant concerns about the fairness and ethical implications of using such AI systems in critical decision-making processes. Addressing these biases is essential to ensure that AI systems do not perpetuate existing inequalities and instead contribute to a fairer and more just society.

\subsection{Fairness Auditing}

One effective approach to achieving fairness in AI systems is to conduct audits. Quantitative fairness auditing involves evaluating these systems using various fairness metrics to identify potential biases and inequalities. Some commonly proposed metrics for assessing AI system fairness include demographic parity, equalized odds, and accuracy equality. These metrics provide a quantitative foundation for addressing fairness concerns in AI models.

A well-documented example of fairness auditing is the investigation of the COMPAS system. Journalists from ProPublica revealed that COMPAS exhibited bias against certain demographic groups, particularly African Americans. The system was more likely to falsely classify African American defendants as high risk for reoffending compared to their white counterparts. This finding not only underscored the necessity of fairness auditing but also highlighted the potential harm of deploying biased AI systems in critical areas.

This case emphasizes the importance of independent audits in ensuring fairness. Evaluations conducted by third parties, such as those performed by ProPublica, are vital because they operate free from the influence of those who develop or deploy the systems. By offering an objective perspective, third-party audits can identify and address biases that internal evaluations might overlook. Without such oversight, AI systems risk perpetuating systemic inequalities, making independent auditing a crucial aspect of accountability in AI development.

@@

\subsection{Challenges in Auditing}

While fairness auditing is essential, it presents significant challenges, particularly when it involves sensitive data. To conduct audits, auditors may rely on access to real datasets containing sensitive information, raising significant security and privacy concerns. Auditors often require access to real-world datasets containing sensitive attributes, such as race, gender, or income, to evaluate an AI system's fairness. However, this reliance on sensitive data raises concerns about data privacy and security.

This reliance invites data security attacks on auditors, making them targets of unauthorized access. The use of real datasets for fairness auditing creates a point of vulnerability, making auditors potential targets for data breaches and unauthorized access. If sensitive data falls into the wrong hands, it could lead to severe consequences, including identity theft, discrimination, and reputational damage.

Moreover, analyses on real data runs the risk of data inference attacks, where confidential information may be deduced from seemingly harmless or aggregated data. Even if the real data is not directly exposed, the results of fairness analyses can sometimes reveal sensitive information through inference attacks. For example, attackers could combine publicly available audit results with external datasets to infer private details about individuals in the original dataset.

\subsection{Differentially Privacy}

Various technologies have been developed to mitigate these risks, such as differential privacy. Differential privacy is a mathematical framework that ensures the inclusion or exclusion of any individual in a dataset has a minimal and quantifiable impact on the overall results of data analysis. This is achieved by introducing carefully calibrated random noise into the computations derived from the data, ensuring that individual records remain indistinguishable while preserving aggregate patterns and trends.

The concept of differential privacy is particularly valuable in scenarios where sensitive information, such as health records or financial data, is involved. For example, a differential privacy mechanism could allow researchers to analyze patterns in patient health outcomes without exposing any single patient's medical history. This approach strikes a balance between data utility and privacy, enabling meaningful analysis while safeguarding individuals against privacy breaches or inference attacks.

The key idea is that no matter how much background knowledge an attacker has about the data, they cannot determine with high confidence whether a specific individual is included in the dataset. This is quantified by some parameters which control the amount of noise introduced. Smaller parameters provide stronger privacy but can reduce the accuracy of the analysis, while larger parameters allow for more precise results but with weaker privacy guarantees. The challenge is to carefully calibrate the parameters to ensure that privacy is maintained without overly compromising the utility of the data.

\subsection{Synthetic Data}

Synthetic data are fake data that is generated from real data. Synthetic data is artificially generated to replicate the statistical properties of real data while protecting individual privacy. The principle of this technology is to preserve some statistical properties of the real data while protecting the privacy of the individuals in the real data. Unlike anonymized data, synthetic data breaks the link to the original data, reducing the risk of re-identification.

Among the generation techniques, differentially private synthetic data offers a way to investigate the data properties under differential privacy guarantees. It stands out for it strong differential privacy guarantees when generating synthetic data. This method combines data synthesis with the principles of differential privacy, ensuring that no single individual's information significantly influences the synthetic dataset. By introducing controlled randomness during generation, it provides strong privacy guarantees while preserving the overall utility of the data.

Differentially private synthetic data is particularly valuable in sensitive domains like healthcare and finance, enabling researchers to analyze patterns or train AI models without exposing personal details. It also supports fairness auditing by allowing the examination of datasets and models without revealing sensitive attributes. This dual focus on privacy and utility makes it a valuable tool.

\subsection{Differentially Private Synthetic Data for Fairness Auditing}

@@

We propose a novel auditing framework that leverages differentially private synthetic data to evaluate the fairness of AI systems. Our framework lets the auditors assess the fairness of AI systems without exposing sensitive information. By using differentially private synthetic data, auditors can conduct fairness analyses without directly exposing sensitive information.

This method offers a unique solution to reconcile the need for fairness auditing with the imperative to protect privacy. By preserving key statistical properties of the original data, synthetic data allows auditors to assess fairness metrics accurately while mitigating the risks of data breaches and privacy violations.

This technology enables auditing without exposing sensitive information, offering a privacy-preserving way to assess AI systems. However, synthetic data also introduces new challenges. One major concern is whether it accurately reflects the fairness properties of the real data. Whether or not it accurately preserves the fairness properties of the real data is a critical question.

Since fairness metrics depend on the relationships between sensitive attributes and outcomes, any discrepancies in how synthetic data mimics these relationships could lead to inaccurate conclusions in fairness audits. Ensuring that synthetic data preserves these properties is critical for reliable auditing.

\subsection{Contribution}

@@

In this paper, we propose a novel framework for fairness auditing that leverages differentially private synthetic data. Our framework enables auditors to evaluate the fairness of AI systems without exposing sensitive information, addressing both privacy and security concerns.

@@

Through our experiments, we evaluate the extent to which synthetic data can approximate fairness metrics computed on original datasets. Our findings contribute to the growing body of research on ethical AI and offer insights for practitioners seeking to ensure fairness in AI systems.

\subsection{Structure}

The rest of this paper is organized as follows. Section 2 provides the background of related work. Section 3 provides the preliminary knowledge. Section 4 outlines the motivation behind our proposed framework. Section 5 describes our methodology for generating differentially private synthetic data and evaluating fairness metrics. Section 6 presents our experimental setup and results. Section 7 discusses the implications of our findings. Section 8 concludes the paper and outlines future work.

% TODO: future work?
